---
title: "MTH6139 Time Series"
subtitle: "Coursework 1 - Forecasting the GBP/USD Exchange Rate with Prophet"
author: "Mohammad Ruhan-Ur Rahman"
date: "Spring Term 2025"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: spacelab
    highlight: tango
---
 
```{r, echo=FALSE}
htmltools::img(src = knitr::image_uri("images/QMlogo.png"),
               alt = 'logo',
               style = 'position:absolute; top:0; right:0; padding:10px; width:20%;')
```

# Introduction

This project analyses the GBP/USD exchange rate using Meta's Prophet model. The aim is to identify trends and produce a short-term forecast.

# Data and setup

The data is daily GBP/USD exchange rates downloaded from Yahoo Finance.

```{r message=FALSE, warning=FALSE}
library(quantmod)
library(prophet)
library(ggplot2)

getSymbols("GBPUSD=X", src = "yahoo", from = "2020-01-01")

gbp_usd_data <- data.frame(
  ds = as.Date(index(`GBPUSD=X`)),
  y = as.numeric(Cl(`GBPUSD=X`))
)
```

# Exploratory analysis

```{r fig.width=10, fig.height=5}
ggplot(gbp_usd_data, aes(x = ds, y = y)) +
  geom_line() +
  labs(title = "GBP/USD Exchange Rate", x = "Date", y = "Rate")
```

The series shows high volatility and no clear seasonal pattern.

```{r}
summary(lm(y ~ ds, data = gbp_usd_data))
```

A simple trend suggests the exchange rate changes over time but is unstable.

# Prophet model

```{r message=FALSE, warning=FALSE}
model <- prophet(gbp_usd_data)

future <- make_future_dataframe(model, periods = 30)

forecast <- predict(model, future)
```

```{r fig.width=10, fig.height=6}
plot(model, forecast)
```

The forecast shows short-term movement with wide uncertainty.

```{r fig.width=10, fig.height=8}
prophet_plot_components(model, forecast)
```

Trend is the main feature, with weak seasonality.

# Alternative model

```{r message=FALSE, warning=FALSE}
model2 <- prophet(gbp_usd_data,
                  weekly.seasonality = FALSE,
                  yearly.seasonality = FALSE)

forecast2 <- predict(model2, make_future_dataframe(model2, periods = 30))
```

```{r fig.width=10, fig.height=6}
plot(model2, forecast2)
```

Removing seasonality does not change results much.

# Discussion

The GBP/USD series is more irregular than standard textbook time series. This means forecasts should be interpreted carefully, especially during periods of market uncertainty.

# Conclusion

GBP/USD shows volatility and weak seasonality. Prophet works well for visualising trends but forecasts should be interpreted carefully.

# References

- Prophet documentation: <https://facebook.github.io/prophet/docs/quick_start.html#r-api>
- Yahoo Finance: <https://finance.yahoo.com>